---
title: "chapter6.Rmd"
output: html_document
theme: cerulean
---
 
# Chapter 6 begins here  {#anchor} 

This week we are learning about longitudinal data. Longitudinal data is the name of data where multiple measurements is done at different times on the same subject. That way, we get multiple responses for each variable. The different measurements are intercorrelated,and the different variables measured can also be correlated to each oter. Hence,we can have two different types of dependecies in the data.
```{r}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

# Access the packages dplyr and tidyr
library(dplyr)
library(tidyr)

# Factor ID and Group
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# Convert to long form and extract day numbers from WD
RATSL <- RATS %>% gather(key = WD, value = Weight, -ID, -Group) %>% mutate(Time = as.integer(substr(WD, 3,4))) 

# Factor treatment & subject
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# Convert to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)

# Extract the week number
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))
```
The first data used this week is the RATS data. It consists of observations of rats' weights and weight-changes during a nine week period. The rats are separated into three different groups, and each group is put on different diets.
In the second data, the BPRS, 40 males have been assigned to two different treatments. 18 symptoms, such as hostility, hallucinations, grandiosity and suspiciousness are being observed in the males weeklyfor 8 weeks. The observations are based on a brief psychiatric rating score (BPRS), which gives each symptom a score from 0 (not present) to seven (extremely severe). This scale helps to evaluate whether a patient suffers from schizofrenia or not.
```{r}
names(RATSL)
str(RATSL)
summary(RATSL)
names(BPRSL)
str(BPRSL)
summary(BPRSL)
```
I will be using the data in the long form, i.e. RATSL and BPRSL

### Part 2: RATS {.css_class} 

###### graphical display 
In the first part of the analysis, I willbe using the RATS data set.
To begin the analysis, I will create a plot of the individual response profiles in RATS. In the first plot the data is non-standardize, and in the second plot the data is standardized.
```{r}
#Access the package ggplot2
library(ggplot2)

# Draw the plot of RATSL
ggplot(RATSL, aes(x = Time, y = Weight, linetype = as.factor(ID))) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```
From the plot, which is made from the non-standardized data on the rats and their wegihts, we can see that there are wegiht differences between the rats. The rats in Group 1 started off at the smallest weight (225-275g), while those in Gruoup two and three were a bit heavier (400-550g).The rats' weight in Group one did not change much during the experiment. Some rats in Group two gained almost 100g, and all rats in Group two gained at least a bit weight (~25g at a minimum). The rats in Group 3 also gained weight, but their weight curves is more uneven than those in Group one and two. I think it would be better to mix the rats, so that the weight distribution between the groups would be more even, since the rats at different weights might have different eating habits, which could affect the results.

Next the data is standardized, and then the same plot is drawn again.
```{r}
# Standardise the variable Weight
RATSLS <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()

# Plot with standardized values
ggplot(RATSLS, aes(x = Time, y = stdWeight, linetype = as.factor(ID))) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")
```
The standardized plot differs from the non-standardized plot. The subjects with the highest values in the beginning tend to have higher values throughtout the study. This is called tracking, and it can be more clearly visualized by standardized plots. 
    The rats' weight-curves in Group two is more even, and one rat(number7) seems to even have lost weight, even though it looked like it had gained a little weight in the non-standardized plot! Almost all rats seem to have lost weight in the third group, except for one rat.The rat that didn't lose weight did not gain wegiht etiher, and it was the one with the lowest initial weight, which could affect the results.
    
    
When dealing with large amounts of data, plots of individual response gets messy and unclear, and is thereforeof little use. A better way of showing progress is to calculate mean profiles for the different groups throughout the study, and then compare them with each other in summary graphs. The standard error of mean can be used to indicate the variation of the observations at different time points.
```{r}
# Time 
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of weight by group and time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = as.factor(Group), shape = as.factor(Group))) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
The standard error is an estimate of the standard deviation. If the standrad erroris big in a group, then the variation is big in that group. This summary plot is quite clear, and it shows that on average the rats in all groups got fatter from the diets. The second group has the highest standard error, and the first group has the lowest standard error. The third group has the highest values (weight) while the first group has the lowest values (weight).

Next I will draw a boxplot of the mean weight of the rats during the study. First I will calculate the mean wegiht in the different groups.
```{r}
#creating a summary data by Group and IDwith the mean as the summary variable
RATSLT <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

#drawing a boxplot of mean vs Group
ggplot(RATSLT, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days")
```
The summary boxplot shows that there is one ID that is outside the box, i.e. it's much higher or lower than the rest of the rats in that group. This phenomenon is called an outlier. The outlier can alter the results of the whole group, since it affects the mean more thanthe average rat in the group. Lets try and leave out the outliers from the mean to see what difference it makes. 
```{r}
RATSLT1 <- filter(RATSLT, (Group ==1 & mean >250) |(Group == 2 & mean < 550) | (Group == 3 & mean >500))
ggplot(RATSLT1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days")
```
After removing the outlie, the variance in the mean is much lower. This mean that the outlier affected the mean of the group more than the rest of the rats affect it. For example in group two where the largest change could be seen, the rest of the rats seem to be in the almost same weight, while the outlier was drastically above that weight (the outlier's weight is approximately 600g and the rest of the rats'meanweight is around 450g). Since the outlier was much heavier than rest, its weight affected the mean more. 


Although the graphs have shown a clear difference between the groups' weights, and therefore a difference in howthediets work, it is still best to test the data. This can be done with a t-test. I will assess the difference between the groups, and calculate a confidence interval for the difference. The data without the outlier will be used, since it gives the most precise data. Since I have three groups, i will use the ANOVA to find out the differences in weight.
```{r}

# Add the baseline from the original data as a new variable to the summary data
RATSLT2 <- RATSLT %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSLT2)
anova(fit)
```

The basline tells us that the significance is great. We can also conclude that the Groupvarible has 95% confidence interval, which means that thatthe diet group that the rats belong to is significant.

### Part 3: BPRS {.css_class}

To begin part 3, I'm going to draw a plot of the data and explore its structure. In the first plot the data's longitudinal nature is ignored.
```{r}
dim(BPRSL)
str(BPRSL)
names(BPRSL)
ggplot(BPRSL, aes(x = week, y = bprs,linetype =as.factor(subject))) +
  geom_line() +scale_linetype_manual(values = rep(1:10, times = 4)) + facet_grid(.~treatment,labeller =label_both) + theme(legend.position = "top") + scale_y_continuous(limits = c(min(BPRSL$bprs),max(BPRSL$bprs)))
```
This is a plot showing the individuals bprs scoresthroughout the study (8weeks). We can see how they change, and how the two treatment groups differ from each other. The bprs scores seem to be declining in most cases, indicating that the treatments are working. There are, however also some individuals where the scores have risen during the treatment weeks. Also, from the plot it seems like the bprs scores from the treatment 1 group is convering in the end. This is quite interesting.

Next I'mgoing to fit a multiple linear regression model with bprs as response variable and week and treatment as explanatory variables
```{r}
# create a regression model bprs_reg
bprs_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(bprs_reg)
```
By using this method, we ignore that the repeated-measured structure of the data and treat it as separate independent measurements. The significance of the variableweek is noteworthy, and treatment 1 and 2 differ from each other.

###### Random Intercept and Random Slope Models
Now we're going to move on to more appropriate graphics and models. First we're going to fit a random intercept model for the explanatory variables week and treatment. This allows for the linear regression for each subject to differ in intercept from the other subjects.
```{r}
library(lme4)
# Create a random intercept model
bprs_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(bprs_ref)
```
The standard deviation of the subject is 6.885,which can be seen from the summary.

Next,I'm going to fit the random intercept and random slope model onthe bprs data. This allows the linearregression fits fro each individualto differ in intercept and also in slope. This way we can account for individual subjects' bprs scores, and also how the change happens over time.

```{r}
# create a random intercept and random slope model
bprs_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(bprs_ref1)

# perform an ANOVA test on the two models
anova(bprs_ref1,bprs_ref)
```
The ANOVA-test tells us that the p value of the likelihood test between intercept and slope is 0.026 and significant in 99% confidence level. The lower the value of the p-value, the better the model fits. This caseis a good fit for the model.


For the final part I'm going to fit a random intercept and slope model on the data. This allows for a group x time interaction. I'm also going to perform a ANOVA-test on the data,to tell how well the model fits.
```{r}
# create a random intercept and random slope model with the interaction
bprs_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(bprs_ref2)

# perform an ANOVA test on the two models
anova(bprs_ref2, bprs_ref1)
```
According to the ANOVA-test,week xtreatment interaction is significant in 95% confidence level
```{r}
# draw the plot of BPRSL
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_linetype_manual(values =rep(1:10, times =4)) +
  scale_y_continuous(name = "bprs", limits = c(min(BPRSL$bprs), max(BPRSL$bprs))) +facet_grid(.~treatment, labeller = label_both) +
  theme(legend.position = "top")
```
```{r}
# Create a vector of the fitted values
Fitted <- fitted(bprs_ref2)

# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>% mutate(Fitted)


# draw the plot of BPRSL with the Fitted values
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) +scale_linetype_manual(values =rep(1:10, times =4)) +
  scale_y_continuous(name = "bprs", limits = c(min(BPRSL$bprs), max(BPRSL$bprs))) +facet_grid(.~treatment, labeller = label_both) +
  theme(legend.position = "none")
``` 
In the second plotI tried to create a vector of fitted values. Then i tried to add the vector as a column in BPRSL and then i tried to create a new plot. This did, however, not work at all, and Idon't see what I did wrong.. Anyhow, without a proper graph it's hard to do a good interpretation on it.