---
title: "chapter4.Rmd"
output: html_document
---

```{r}
# access the MASS package
library(MASS)

# load the data
data("Boston")


str(Boston)
dim(Boston)
```
The Boston dataset describes the prices of houses in different parts of Boston. It does also give data about other things related to the housing, such as crime, age distribution among the house owners and areas of non-retail busísness in Boston. there are 14 variables in total, and 506 observations.

```{r}
library("corrplot")
pairs(Boston)
summary(Boston) 
cor_matrix<-cor(Boston)
corrplot(cor_matrix, method = "circle")
```
The corrplot gives an overview of the correlations between the variables. For example, there seems to be a fairly strong positive correlation between taxes and distance to radial highways. A positive correlation indicates that when one variable increases the other does as well. Furthermore there seems to be a pretty strong negative correlation between nitrogen oxides concentration and distance to Boston employment centres. This means that the nitrogen oxides concentrations decrease as the distance from the employment centres increase. The blue circles means positive correlations and the read means negative correlation. The bigger the circle is, the stronger the correlations is.

```{r}
boston_scaled <- scale(Boston)
summary(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)
```
After scaling the dataset, the values become much lower. Many of the minimum values even become negative. This makes it possible to compare the results later, since it would be hard to compare variables with widely different scales.
```{r}
bins <- quantile(boston_scaled$crim)
vector <- c("low", "med_low", "med_high", "high")
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE,label = vector)
boston_scaled <- dplyr::select(boston_scaled, -crim)
boston_scaled <- data.frame(boston_scaled, crime)
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

#fitting the lda on the train set
lda.fit <- lda(formula = crime ~. , data = train)
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "purple", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}
classes <- as.numeric(train$crime)
plot(lda.fit, dimen = 2, col =classes, pch=classes)
lda.arrows(lda.fit, myscale = 4)
```
the LDA shows the relationship between the different variables and crime in Boston. It shows how different things seem to be connected to crime levels. The same things can be read from the corrplot earlier by comparing different variables with the crime variable. The arrow going out of the picture is the "rad" arrow.
```{r}
# save the correct classes from test data

correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)
pred <- lda.pred$class

``` 
According to the cross tabulation, there were no false predictions in the high crime predictions. There we're a few (5) mispredictions in the med_high column, a few more mispredictions in the med_low column and some mispredictions in the low crime column. 